---
title: "Differentially Open Chromatin Region Analysis"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
This document describes analysis of peaks for differentially open chromatin regions. The analysis was performed using DESeq2 package. 

```{r packages, include=FALSE}
#Install and load required R packages. Uncomment the below install lines if you do not already have them installed
#BiocManager::install("DESeq2")
#BiocManager::install("apeglm")
#library(ChIPseeker)
library(DESeq2)
library(ggvenn)
library(ggVennDiagram)
library("gplots")
library("heatmaply")
```

## Load featurecounts data (example rows below)
```{r loadData, echo=FALSE, warning=FALSE, message=FALSE}
## Load featurecount data and sample information data
cts <- as.matrix(read.csv("tko_cmt_pbs_wt_featureCounts_for_deseq2.txt",sep="\t",row.names="Geneid"))
head(cts)
```

## Load sample information data
```{r loadsamplinfo, echo=FALSE, warning=FALSE, message=FALSE}
## Load sampleinfo file
coldata <- read.csv("sampleinfo.txt",sep="\t",row.names="ID")
coldata  
```

## Differential Analysis
Differential analysis is done with DESeq2, after removing rows with a total count of less than 10, an alpha of 0.05 and pvalue adjustment using BH method.

```{r differentialanalysis, echo=FALSE, warning=FALSE, message=FALSE}
## Convert count matrix to DESeq data set, with Sample column as design factor
dds <- DESeqDataSetFromMatrix(countData = cts,colData = coldata,design = ~ Sample)

## Filter out low count peaks
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

## Set the contrast reference levels
dds$Sample <- relevel(dds$Sample, ref = "TKO_Cmst")
dds_TKO_Cmst <- DESeq(dds)

dds$Sample <- relevel(dds$Sample, ref = "WT_Cmst")
dds_WT_Cmst <- DESeq(dds)
```

### TKO_PBS vs TKO_Cmst
```{r tkopbscmst, echo=FALSE, warning=FALSE, message=FALSE}
## Generate deseq2 differential results objects
res_tko_pbs_cmst <- results(dds_TKO_Cmst, contrast=c("Sample","TKO_PBS","TKO_Cmst"), alpha=0.05, pAdjustMethod="BH")
summary(res_tko_pbs_cmst)
```

### WT_PBS vs WT_Cmst
```{r wtpbscmst, echo=FALSE, warning=FALSE, message=FALSE}
res_wt_pbs_cmst <- results(dds_WT_Cmst, contrast=c("Sample","WT_PBS","WT_Cmst"), alpha=0.05, pAdjustMethod="BH")
summary(res_wt_pbs_cmst)
```

### WT_Cmst vs TKO_Cmst
```{r wtcmsttkocmst, echo=FALSE, warning=FALSE, message=FALSE}
res_cmst_wt_tko <- results(dds_TKO_Cmst, contrast=c("Sample","WT_Cmst","TKO_Cmst"), alpha=0.05, pAdjustMethod="BH")
summary(res_cmst_wt_tko)
```

```{r showcontrast, include=FALSE}
## Show the contrast - coef's
resultsNames(dds_TKO_Cmst)
resultsNames(dds_WT_Cmst)
```

```{r shrinklfc, include=FALSE}
## Shrink pvalues - https://doi.org/10.1093/bioinformatics/bty895
resLFC_tko_pbs_cmst <- lfcShrink(dds_TKO_Cmst, coef="Sample_TKO_PBS_vs_TKO_Cmst", type="apeglm", res=res_tko_pbs_cmst)
resLFC_wt_pbs_cmst <- lfcShrink(dds_WT_Cmst, coef="Sample_WT_PBS_vs_WT_Cmst", type="apeglm", res=res_wt_pbs_cmst)
resLFC_cmst_wt_tko <- lfcShrink(dds_TKO_Cmst, coef="Sample_WT_Cmst_vs_TKO_Cmst", type="apeglm", res=res_cmst_wt_tko)

## Open an empty svg device
#png(filename="maplot-all-diffpeaks.png", width=8, height=3)
```

## MA Plots (differential data for all data points)
```{r maplots, echo=FALSE, warning=FALSE, message=FALSE}
## Set facet panels and plot
par(mfrow=c(1,3), mar=c(4,4,2,1))
xlim <- c(1,1e5); ylim <- c(-3,3)
plotMA(resLFC_tko_pbs_cmst, xlim=xlim, ylim=ylim, main="TKO_PBS vs TKO_Cmst")
plotMA(resLFC_wt_pbs_cmst, xlim=xlim, ylim=ylim, main="WT_PBS vs WT_Cmst")
plotMA(resLFC_cmst_wt_tko, xlim=xlim, ylim=ylim, main="WT_Cmst vs TKO_Cmst")
#run dev.off() if figure margin is too large
#dev.off()
```
<br>
## Significanly differentially open chromatin regions (OCR's)
DESeq2 results are ordered by pvalue, filtered for a padj < 0.05 and log2fold change >=1.<br><br>

### Column descriptions
```{r columndescription, echo=FALSE, warning=FALSE, message=FALSE}
## Describe column headers for the differetnially open chromatin region files
mcols(resLFC_tko_pbs_cmst)$description
```
### Full DESeq2 results are saved in csv files<br>
<a href="TKO_PBS_vs_TKO_Cmst_full_results.csv">TKO_PBS_vs_TKO_Cmst_full_results.csv</a><br>
<a href="WT_PBS_vs_WT_Cmst_full_results.csv">WT_PBS_vs_WT_Cmst_full_results.csv</a><br>
<a href="WT_Cmst_vs_TKO_Cmst_full_results.csv">WT_Cmst_vs_TKO_Cmst_full_results.csv</a><br>

```{r savingfull, echo=FALSE, warning=FALSE, message=FALSE}
## Sort results based on pvalue column
resOrderedLFC_tko_pbs_cmst <- resLFC_tko_pbs_cmst[order(resLFC_tko_pbs_cmst$pvalue),]
resOrderedLFC_wt_pbs_cmst <- resLFC_wt_pbs_cmst[order(resLFC_wt_pbs_cmst$pvalue),]
resOrderedLFC_cmst_wt_tko <- resLFC_cmst_wt_tko[order(resLFC_cmst_wt_tko$pvalue),]

## Save sorted full results to .csv files
write.csv(as.data.frame(resOrderedLFC_tko_pbs_cmst), file="TKO_PBS_vs_TKO_Cmst_full_results.csv")
write.csv(as.data.frame(resOrderedLFC_wt_pbs_cmst), file="WT_PBS_vs_WT_Cmst_full_results.csv")
write.csv(as.data.frame(resOrderedLFC_cmst_wt_tko), file="WT_Cmst_vs_TKO_Cmst_full_results.csv")
```

### Significantly differentially OCR's lists are saved in csv files<br>
<a href="TKO_PBS_vs_TKO_Cmst_sig_results.csv">TKO_PBS_vs_TKO_Cmst_sig_results.csv</a><br>
<a href="WT_PBS_vs_WT_Cmst_sig_results.csv">WT_PBS_vs_WT_Cmst_sig_results.csv</a><br>
<a href="WT_Cmst_vs_TKO_Cmst_sig_results.csv">WT_Cmst_vs_TKO_Cmst_sig_results.csv</a><br>

```{r savesiglist, echo=FALSE, warning=FALSE, message=FALSE}
## Filter for significantly differentially open chromatin regions - padj 0.05, logfold 1 (fold 2)
resSigOrderedLFC_tko_pbs_cmst <- subset(resOrderedLFC_tko_pbs_cmst, padj < 0.05 & abs(log2FoldChange) >= 1)
resSigOrderedLFC_wt_pbs_cmst <- subset(resOrderedLFC_wt_pbs_cmst, padj < 0.05 & abs(log2FoldChange) >= 1)
resSigOrderedLFC_cmst_wt_tko <- subset(resOrderedLFC_cmst_wt_tko, padj < 0.05 & abs(log2FoldChange) >= 1)

## Save significantly differentially open chromatin regions to .csv files
write.csv(as.data.frame(resSigOrderedLFC_tko_pbs_cmst), file="TKO_PBS_vs_TKO_Cmst_sig_results.csv")
write.csv(as.data.frame(resSigOrderedLFC_wt_pbs_cmst), file="WT_PBS_vs_WT_Cmst_sig_results.csv")
write.csv(as.data.frame(resSigOrderedLFC_cmst_wt_tko), file="WT_Cmst_vs_TKO_Cmst_sig_results.csv")
```

## Extract individual sample data for significant OCR's
Save normalized individual sample data in csv files <br>
<a href="TKO_PBS_vs_TKO_Cmst_sig_normalized_data.csv">TKO_PBS_vs_TKO_Cmst_sig_normalized_data.csv</a><br>
<a href="WT_PBS_vs_WT_Cmst_sig_normalized_data.csv">WT_PBS_vs_WT_Cmst_sig_normalized_data.csv</a><br>
<a href="WT_Cmst_vs_TKO_Cmst_sig_normalized_data.csv">WT_Cmst_vs_TKO_Cmst_sig_normalized_data.csv</a><br>

```{r extractsampledata, echo=FALSE, warning=FALSE, message=FALSE}
## Store the row names of significant peaks
summary(selectedresSigOrderedLFC_tko_pbs_cmst<-rownames(resSigOrderedLFC_tko_pbs_cmst))
summary(selectedresSigOrderedLFC_wt_pbs_cmst<-rownames(resSigOrderedLFC_wt_pbs_cmst))
summary(selectedresSigOrderedLFC_cmst_wt_tko<-rownames(resSigOrderedLFC_cmst_wt_tko))

## Retrieve and store normalized data for all samples, for the significant peaks
selectedtko_pbs_cmst = log2(counts(dds_TKO_Cmst,normalized=TRUE)[rownames(dds_TKO_Cmst) %in% selectedresSigOrderedLFC_tko_pbs_cmst,])
selectedwt_pbs_cmst = log2(counts(dds_WT_Cmst,normalized=TRUE)[rownames(dds_WT_Cmst) %in% selectedresSigOrderedLFC_wt_pbs_cmst,])
selectedcmst_wt_tko = log2(counts(dds_TKO_Cmst,normalized=TRUE)[rownames(dds_TKO_Cmst) %in% selectedresSigOrderedLFC_cmst_wt_tko,])

## Save normalized data for all samples, for the significant peaks to .csv files
write.csv(as.data.frame(selectedtko_pbs_cmst), file="TKO_PBS_vs_TKO_Cmst_sig_normalized_data.csv")
write.csv(as.data.frame(selectedwt_pbs_cmst), file="WT_PBS_vs_WT_Cmst_sig_normalized_data.csv")
write.csv(as.data.frame(selectedcmst_wt_tko), file="WT_Cmst_vs_TKO_Cmst_sig_normalized_data.csv")
```

## Clustered heatmaps
### TKO_PBS vs TKO_Cmst
```{r heatmap1, echo=FALSE, warning=FALSE, message=FALSE}
heatmap.2(log2(counts(dds_TKO_Cmst,normalized=TRUE)[rownames(dds_TKO_Cmst) %in% selectedresSigOrderedLFC_tko_pbs_cmst,]), main="TKO_PBS vs TKO_Cmst", scale="row", trace="none", col="heat.colors", labRow=FALSE,cexRow=1,cexCol=1,margins=c(12,8))
```

### WT_PBS vs WT_Cmst
```{r heatmap2, echo=FALSE, warning=FALSE, message=FALSE}
heatmap.2(log2(counts(dds_WT_Cmst,normalized=TRUE)[rownames(dds_WT_Cmst) %in% selectedresSigOrderedLFC_wt_pbs_cmst,]), main="WT_PBS vs WT_Cmst", scale="row", trace="none", col="heat.colors", labRow=FALSE,cexRow=1,cexCol=1,margins=c(12,8))
```

### WT_Cmst vs TKO_Cmst
```{r heatmap3, echo=FALSE, warning=FALSE, message=FALSE}
heatmap.2(log2(counts(dds_TKO_Cmst,normalized=TRUE)[rownames(dds_TKO_Cmst) %in% selectedresSigOrderedLFC_cmst_wt_tko,]), main="WT_Cmst vs TKO_Cmst", scale="row", trace="none", col="heat.colors", labRow=FALSE,cexRow=1,cexCol=1,margins=c(12,8))
```

```{r alternateheatmaps, echo=FALSE, warning=FALSE, message=FALSE}
#heatmaply(selectedtko_pbs_cmst, k_col=2, k_row=7, col=rainbow, trace="none", column_text_angle=90, showticklabels = c(T,F),main="TKO_PBS vs TKO_Cmst")
#heatmaply(selectedwt_pbs_cmst, k_col=2, k_row=7, col=rainbow, trace="none", column_text_angle=90, showticklabels = c(T,F),main="WT_PBS vs WT_Cmst")
#heatmaply(selectedcmst_wt_tko, k_col=2, k_row=7, col=rainbow, trace="none", column_text_angle=90, showticklabels = c(T,F),main="WT_Cmst vs TKO_Cmst")

#ht1<-Heatmap(log2(counts(dds_TKO_Cmst,normalized=TRUE)[rownames(dds_TKO_Cmst) %in% selectedresSigOrderedLFC_tko_pbs_cmst,]), show_row_names = FALSE, column_title = "TKO_PBS vs TKO_Cmst")
#ht2<-Heatmap(log2(counts(dds_WT_Cmst,normalized=TRUE)[rownames(dds_WT_Cmst) %in% selectedresSigOrderedLFC_wt_pbs_cmst,]), show_row_names = FALSE, column_title = "WT_PBS vs WT_Cmst")
#ht3<-Heatmap(log2(counts(dds_TKO_Cmst,normalized=TRUE)[rownames(dds_TKO_Cmst) %in% selectedresSigOrderedLFC_cmst_wt_tko,]), show_row_names = FALSE, column_title = "WT_Cmst vs TKO_Cmst")
```

## Significant Peak Comparison
The below venndiagram compares significant OCR's between TKO_PBS vs TKO_Cmst, WT_PBS vs WT_Cmst, WT_Cmst vs TKO_Cmst.<br>

```{r ocrVennDiagram, echo=FALSE, warning=FALSE, message=FALSE}
# Plot venndiagram between comparison ocrs
# significant ocr's list for all comparisons

ocrs<-list(TKOPBS_TKOCmst=selectedresSigOrderedLFC_tko_pbs_cmst, WTPBS_WTCmst=selectedresSigOrderedLFC_wt_pbs_cmst, WTCmst_TKOCmst=selectedresSigOrderedLFC_cmst_wt_tko)

# Generate interactive venndiagram of ocr overlap between comparisons
ggVennDiagram(ocrs, label = c("count"))

```

## Save R session
```{r savesession}
save.image("deseq2.RData")
```

## Summary
The Cmst has the most effect along with the mutant, showing 4816 significantly OCR's in TKO_PBS vs TKO_Cmst (2330 unique - as seen in the venndiagram).<br>
WT_PBS vs WT_Cmst has 3524 significanly OCR's<br>
The mutant alone (WT_Cmst vs TKO_Cmst) has the least number (680) of significantly OCR's<br> 

## Reference
1. DESeq2: https://bioc.ism.ac.jp/packages/3.7/bioc/html/DESeq2.html<br>
2. LFCShrink: https://doi.org/10.1093/bioinformatics/bty895 <br>
3. HeatMap2: https://earlglynn.github.io/RNotes/package/gplots/heatmap2.html <br>

## R Session Info
```{r sessionfo}
sessionInfo()
```
