---
title: "Enrichment Analysis of Differentially Open Chromatin Regions"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
This document describes KEGG pathway and Ontology enrichment analysis of differentially open chromatin regions. The analysis was performed using the significantly differentially OCR results from DESeq2 package

```{r packages, include=FALSE}
#Install and load required R packages. Uncomment the below install lines if you do not already have them installed
library(tidyverse)
library(tidyr)
library(dplyr)
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
BiocManager::install("ReactomePA")
library(ReactomePA)
library(clusterProfiler)
library(rmarkdown)
library(gridExtra)
```

```{r datacleaning, include=FALSE}
# Convert deseq2 result to granges format for chipseeker annotation
#tkopbscmst <- as.data.frame(resSigOrderedLFC_tko_pbs_cmst)
#wt_pbs_cmst <- as.data.frame(resSigOrderedLFC_wt_pbs_cmst)
#cmst_wt_tko <- as.data.frame(resSigOrderedLFC_cmst_wt_tko)

# Read differential OCR lists
tkopbscmst <- read.csv("TKO_PBS_vs_TKO_Cmst_sig_results.csv")
wt_pbs_cmst <- read.csv("WT_PBS_vs_WT_Cmst_sig_results.csv")
cmst_wt_tko <- read.csv("WT_Cmst_vs_TKO_Cmst_sig_results.csv")

# Set the column header
names(tkopbscmst)[1] <- "peakName"
names(wt_pbs_cmst)[1] <- "peakName"
names(cmst_wt_tko)[1] <- "peakName"

# Split first column to coordinate parts
tkopbscmst <- tkopbscmst %>% separate(peakName,c("Chr","Start","End"))
wt_pbs_cmst <- wt_pbs_cmst %>% separate(peakName,c("Chr","Start","End"))
cmst_wt_tko <- cmst_wt_tko %>% separate(peakName,c("Chr","Start","End"))

# Add peakname column
res1 = mutate(tkopbscmst, concated_column = paste(Chr, Start, End, sep = '.'))
res2 = mutate(wt_pbs_cmst, concated_column = paste(Chr, Start, End, sep = '_'))
res3 = mutate(cmst_wt_tko, concated_column = paste(Chr, Start, End, sep = '_'))

res1 = res1 %>% relocate(concated_column, .after = End)
res2 = res2 %>% relocate(concated_column, .after = End)
res3 = res3 %>% relocate(concated_column, .after = End)

# Create granges object
myRanges1<- GRanges(seqnames=res1$Chr,ranges=IRanges(as.numeric(res1$Start),as.numeric(res1$End)),strand=NULL,res1$concated_column,res1$baseMean, res1$log2FoldChange,res1$lfcSE,res1$pvalue,res1$padj)
myRanges2<- GRanges(seqnames=res2$Chr,ranges=IRanges(as.numeric(res2$Start),as.numeric(res2$End)),strand=NULL,res2$concated_column,res2$baseMean, res2$log2FoldChange,res2$lfcSE,res2$pvalue,res2$padj)
myRanges3<- GRanges(seqnames=res3$Chr,ranges=IRanges(as.numeric(res3$Start),as.numeric(res3$End)),strand=NULL,res3$concated_column,res3$baseMean, res3$log2FoldChange,res3$lfcSE,res3$pvalue,res3$padj)

# Annotate OCR's
tkopbscmst_annot<-annotatePeak(myRanges1, TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene, annoDb = "org.Mm.eg.db")
wt_pbs_cmst_annot<-annotatePeak(myRanges2, TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene, annoDb = "org.Mm.eg.db")
cmst_wt_tko_annot<-annotatePeak(myRanges3, TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene, annoDb = "org.Mm.eg.db")

tkopbscmst_annot
wt_pbs_cmst_annot
cmst_wt_tko_annot
```

## Save significantly differentially open chromatin regions, with annotations to .csv files
Annotated lists of significantly differentially OCR's are saved in the following files;<br>
<a href="TKO_PBS_vs_TKO_Cmst_sig_results_annotated.csv">TKO_PBS_vs_TKO_Cmst_sig_results_annotated.csv</a><br>
<a href="WT_PBS_vs_WT_Cmst_sig_results_annotated.csv">WT_PBS_vs_WT_Cmst_sig_results_annotated.csv</a><br>
<a href="WT_Cmst_vs_TKO_Cmst_sig_results_annotated.csv">WT_Cmst_vs_TKO_Cmst_sig_results_annotated.csv</a><br>

```{r savedata, include=FALSE}
write.csv(as.data.frame(tkopbscmst_annot), file="TKO_PBS_vs_TKO_Cmst_sig_results_annotated.csv")
write.csv(as.data.frame(wt_pbs_cmst_annot), file="WT_PBS_vs_WT_Cmst_sig_results_annotated.csv")
write.csv(as.data.frame(cmst_wt_tko_annot), file="WT_Cmst_vs_TKO_Cmst_sig_results_annotated.csv")

## Create 6 column BED file for HOMER motif calling
res1 = mutate(res1, strand=".")
res2 = mutate(res2, strand=".")
res3 = mutate(res3, strand=".")

res1 = mutate(res1, Empty="")
res2 = mutate(res2, Empty="")
res3 = mutate(res3, Empty="")

res1 = res1 %>% relocate(strand, .after = concated_column)
res2 = res2 %>% relocate(strand, .after = concated_column)
res3 = res3 %>% relocate(strand, .after = concated_column)

res1 = res1 %>% relocate(Empty, .after = concated_column)
res2 = res2 %>% relocate(Empty, .after = concated_column)
res3 = res3 %>% relocate(Empty, .after = concated_column)

## Save significantly differentially open chromatin regions, in bed format, for homer motif analysis in tab format
write.table(as.data.frame(res1), file="TKO_PBS_vs_TKO_Cmst_sig_results_homer_input.tab", quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")
write.table(as.data.frame(res2), file="WT_PBS_vs_WT_Cmst_sig_results_homer_input.tab", quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")
write.table(as.data.frame(res3), file="WT_Cmst_vs_TKO_Cmst_sig_results_homer_input.tab", quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")

```

### TKO_PBS_vs_TKO_Cmst_sig_results_annotated
```{r displayAnnotations1, echo=FALSE, warning=FALSE, message=FALSE}
#The below set of R code is used to dispaly the annotated OCR's
paged_table(as.data.frame(tkopbscmst_annot))
```
### WT_PBS_vs_WT_Cmst_sig_results_annotated
```{r displayAnnotations2, echo=FALSE, warning=FALSE, message=FALSE}
#The below set of R code is used to dispaly the annotated OCR's
paged_table(as.data.frame(wt_pbs_cmst_annot))
```
### WT_Cmst_vs_TKO_Cmst_sig_results_annotated
```{r displayAnnotations3, echo=FALSE, warning=FALSE, message=FALSE}
#The below set of R code is used to dispaly the annotated OCR's
paged_table(as.data.frame(cmst_wt_tko_annot))
```

## KEGG Pathway Enrichment
No pathway is enriched in WT_Cmst_vs_TKO_Cmst, with an adjusted pvalue cutoff of 0.05

```{r kegg, echo=FALSE, warning=FALSE, message=FALSE}
#pathway1 <- enrichPathway(tkopbscmst_annotdf$geneId, organism = "mouse",pAdjustMethod="none", pvalueCutoff=0.05)
#pathway2 <- enrichPathway(wt_pbs_cmst_annotdf$geneId, organism = "mouse",pAdjustMethod="none", pvalueCutoff=0.05)
#pathway3 <- enrichPathway(cmst_wt_tko_annotdf$geneId, organism = "mouse",pAdjustMethod="none", pvalueCutoff=0.9)

tkopbscmst_annotdf <- as.data.frame(tkopbscmst_annot)
wt_pbs_cmst_annotdf <- as.data.frame(wt_pbs_cmst_annot)
cmst_wt_tko_annotdf <- as.data.frame(cmst_wt_tko_annot)

# Generate kegg pathway enrichments
pathway1 <- enrichKEGG(tkopbscmst_annotdf$geneId, organism = "mouse",pAdjustMethod = "BH", qvalueCutoff = 0.05)
pathway2 <- enrichKEGG(wt_pbs_cmst_annotdf$geneId, organism = "mouse",pAdjustMethod = "BH", qvalueCutoff = 0.05)
pathway3 <- enrichKEGG(cmst_wt_tko_annotdf$geneId, organism = "mouse",pAdjustMethod = "BH", qvalueCutoff = 0.05)

# Plot kegg pathway enrichments
dotplot(pathway1, title="TKO_PBS_vs_TKO_Cmst", orderBy="x")
dotplot(pathway2, title="WT_PBS_vs_WT_Cmst", orderBy="x")
#dotplot(pathway3, title="WT_Cmst_vs_TKO_Cmst", orderBy="x")

```

## Ontology Biological Process Enrichment
No GO BP is enriched in WT_Cmst_vs_TKO_Cmst, with an adjusted pvalue cutoff of 0.05

```{r gobp, echo=FALSE, warning=FALSE, message=FALSE}

# Generate GO Biological Process enrichments
go1 <- enrichGO(gene = tkopbscmst_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "BP",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)
go2 <- enrichGO(gene = wt_pbs_cmst_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "BP",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)
go3 <- enrichGO(gene = cmst_wt_tko_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "BP",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)

# Plot GO BP enrichments
dotplot(go1, title="TKO_PBS_vs_TKO_Cmst", orderBy="x")
dotplot(go2, title="WT_PBS_vs_WT_Cmst", orderBy="x")
#dotplot(go3, title="WT_Cmst_vs_TKO_Cmst", orderBy="x")

#grid.arrange(gop1, gop2, nrow=1)
```

## Ontology Molecular Function Enrichment
No GO MF is enriched in WT_Cmst_vs_TKO_Cmst, with an adjusted pvalue cutoff of 0.05

```{r gomf, echo=FALSE, warning=FALSE, message=FALSE}
# Generate GO MF enrichments
go1m <- enrichGO(gene = tkopbscmst_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "MF",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)
go2m <- enrichGO(gene = wt_pbs_cmst_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "MF",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)
go3m <- enrichGO(gene = cmst_wt_tko_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "MF",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)

# Plot GO MF enrichments
dotplot(go1m, title="TKO_PBS_vs_TKO_Cmst", orderBy="x")
dotplot(go2m, title="WT_PBS_vs_WT_Cmst", orderBy="x")
#dotplot(go3m, title="WT_Cmst_vs_TKO_Cmst", orderBy="x")

#grid.arrange(gop1m, gop2m, nrow=1)
```

## Ontology Cellular Component Enrichment
No GO CC is enriched in WT_Cmst_vs_TKO_Cmst, with an adjusted pvalue cutoff of 0.05

```{r gocc, echo=FALSE, warning=FALSE, message=FALSE}
# Generate GO CC enrichments
go1c <- enrichGO(gene = tkopbscmst_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "CC",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)
go2c <- enrichGO(gene = wt_pbs_cmst_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "CC",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)
go3c <- enrichGO(gene = cmst_wt_tko_annotdf$geneId,keyType = "ENTREZID",OrgDb = org.Mm.eg.db,ont = "CC",pAdjustMethod = "BH",qvalueCutoff = 0.05, readable = TRUE)

## Plot GO CC enrichments
dotplot(go1c, title="TKO_PBS_vs_TKO_Cmst", orderBy="x")
dotplot(go2c, title="WT_PBS_vs_WT_Cmst", orderBy="x")
#dotplot(go3c, title="WT_Cmst_vs_TKO_Cmst", orderBy="x")

#grid.arrange(gop1c, gop2c, nrow=1)
```

## Summary
The significantly differentially OCR's in WT_Cmst_vs_TKO_Cmst doesnt have any pathway or ontologies enriched<br>

## Reference
https://www.rdocumentation.org/packages/clusterProfiler/versions/3.0.4<br>
https://www.bioconductor.org/packages/devel/bioc/vignettes/ChIPseeker/inst/doc/ChIPseeker.html<br>

## R Session Info
```{r sessionfo}
sessionInfo()
```